---
title: "Census_Income/insurance/Education"
author: "VI Nguyen"
date: "2/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
tract_2019 <- read.csv('tract_2019.csv', header = TRUE)
```

```{r}
head(tract_2019)
```
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(modelr))
```

```{r}
unique(tract_2019$State_name)
```


MODEL: FEED THE PERCENTAGES IN THE MODEL AND DO SEE IF THERE IS ANY CORRELATION BETWEEEN THE RESPONSE RATE AND THE PERCENT OF POPULATION
% VS %

MINIMIZE THE CONFOUNDING FACTORS BY RUNNING THE MODEL ON ONE FACTOR AT A TIME
--> SOMETHING DOABLE
--> LANGUAGW: LOW RESPONSE RATE IN DENSE CULTURAL POPULATION
--> AWK
READ IN THE
support vector machine
PCA - great for feature reduction and suspected correlated variables

```{r}
dmv_tract <- tract_2019 %>%
  filter(State_name == 'Virginia' | State_name == 'Maryland' | State_name == 'District of Columbia')
names(dmv_tract)
```

Find column index:

ind <- match(c('Low_Response_Score'), names(dmv_tract))
ind

```{r}
## Extract necessary columns

dmv_others <- dmv_tract %>%
  select( c((3), (468), (220:223), (105), (377),(139), (379:384), (371:373), (284)))
  ## Column 3 is State_name
dmv_others
```


```{r}
## Remove all Margin of Error (MOE) columns

dmv_others <- dmv_others %>%
  select(-contains('MOE'))

head(dmv_others)
```

```{r}
dmv_tract %>%
  filter(pct_Prs_Blw_Pov_Lev_ACS_13_17 > 98)
```

```{r}
## Rate of college students vs. Response rate

dmv_others  %>%
  ggplot() +
  geom_smooth(mapping = aes(x = pct_College_ACS_13_17, y = Low_Response_Score, color = State_name), se = TRUE)
```

```{r}
dmv_others %>%
  ggplot(mapping = aes(x = pct_College_ACS_13_17, y = Low_Response_Score, color = State_name)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_smooth(color = 'black', se = FALSE) + 
  facet_grid(.~State_name)
```

```{r}
dc <- dmv_others %>%
  filter(State_name == 'District of Columbia')
cor.test(dc$pct_College_ACS_13_17, dc$Low_Response_Score)
```

```{r}
dmv_others %>%
  filter(State_name == 'Virginia') %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = "", y = pct_College_ACS_13_17)) + 
  coord_flip() 
```

## Sampling

```{r}
count <- dmv_others %>%
  group_by(State_name) %>%
  summarize(count = n())
count
```

```{r}
## Sample size of 2/3 the population

## DC
sample_size_dc <- floor(.68 * count$count[1])

## Maryland
sample_size_md <- floor(.68 * count$count[2])

## Virginia
sample_size_va <- floor(.68 * count$count[3])
```

```{r}
## Create indeces
sample_dc_ind <- sample(1:count$count[1], sample_size_dc)
sample_md_ind <- sample(1:count$count[2], sample_size_md)
sample_va_ind <- sample(1:count$count[3], sample_size_va)

## Generate random samples
sample_va <- filter(dmv_others, State_name == 'Virginia')[sample_va_ind,]
sample_md <- filter(dmv_others, State_name == 'Maryland')[sample_md_ind,]
sample_dc <- filter(dmv_others, State_name == 'District of Columbia')[sample_dc_ind,]
```

```{r}
sample_va %>%
  ggplot(mapping = aes(x = pct_College_ACS_13_17, y = Low_Response_Score, alpha = 0.5)) +
  geom_point() +
  labs(title = "VA") +
  geom_smooth(method = "lm")
cor.test(sample_va$pct_College_ACS_13_17, sample_va$Low_Response_Score)


sample_md %>%
  ggplot(mapping = aes(x = pct_College_ACS_13_17, y = Low_Response_Score, alpha = 0.5)) +
  geom_point() +
  labs(title = "MD" )+
  geom_smooth(method = "lm")
cor.test(sample_md$pct_College_ACS_13_17, sample_md$Low_Response_Score)


sample_dc %>%
  ggplot(mapping = aes(x = pct_College_ACS_13_17, y = Low_Response_Score)) +
  geom_point() +
  labs(title = "DC") +
  geom_smooth(method = "lm")
cor.test(sample_dc$pct_College_ACS_13_17, sample_dc$Low_Response_Score)
```
```{r}
## Model relationship between % of people in college and low response score

college_model <- lm(Low_Response_Score ~ pct_College_ACS_13_17, data = dmv_others)
college_model%>%
  glance()
```

################################################################################################################################################

## Income

```{r}
## Model relationship between income and low response score

library(modelr)
library(broom)

income_model <- lm(Low_Response_Score ~ Med_HHD_Inc_ACS_13_17, data = dmv_others)
income_model %>%
  glance()
```

```{r}
dmv_others <- dmv_tract %>%
  select( c((3), (468), (220:223), (105), (377),(139), (379:384), (371:373), (284)))
  ## Column 3 is State_name
dmv_others <- dmv_others %>%
  select(-contains('MOE'))
head(dmv_others)
```

```{r}
## Remove the dollar signs

require(stringr)
 ## head(str_sub(as.character(dmv_others$Med_HHD_Inc_ACS_13_17), 2, str_length(as.character(dmv_others$Med_HHD_Inc_ACS_13_17))))
dmv_others$Med_HHD_Inc_ACS_13_17 <- str_sub(as.character(dmv_others$Med_HHD_Inc_ACS_13_17), 2, str_length(as.character(dmv_others$Med_HHD_Inc_ACS_13_17)))
head(dmv_others)
```

```{r}
## Group income into low, lower middle, middle, upper middle, upper clases

dmv_others <- dmv_others %>%
  mutate(Income_level = 
  case_when(as.numeric(gsub(",", "", Med_HHD_Inc_ACS_13_17)) < 31000 ~ "Low income",
            between(as.numeric(gsub(",", "", Med_HHD_Inc_ACS_13_17)), 31000, 42000) ~ "Lower Middle class",
           between(as.numeric(gsub(",", "", Med_HHD_Inc_ACS_13_17)), 42000, 126000) ~ "Middle class",
            between(as.numeric(gsub(",", "", Med_HHD_Inc_ACS_13_17)), 126000, 188000) ~ "Upper Middle class",
           as.numeric(gsub(",", "", Med_HHD_Inc_ACS_13_17)) > 188000 ~ "Upper class"
            
))
dmv_others

```

```{r}
library(forcats)
dmv_others$Income_level <- fct_relevel(dmv_others$Income_level, "Low income", "Lower Middle class", "Middle class", "Upper Middle class")
dmv_others %>%
  ggplot() +
  geom_point(mapping = aes(x = Med_HHD_Inc_ACS_13_17, y = Low_Response_Score, color = Income_level)) 
```

```{r}
library(forcats)
dmv_others$Income_level <- fct_relevel(dmv_others$Income_level, "Low income", "Lower Middle class", "Middle class", "Upper Middle class")
dmv_others%>%
  ggplot() + 
  geom_boxplot(mapping = aes(y = Low_Response_Score , x = Income_level))
```

```{r}
write_csv(dmv_others, path ="dmv_others.csv")
```

```{r}
cov_income_model <- aov(dmv_others$Low_Response_Score ~ dmv_others$Income_level)
cov_income_model
```

```{r}
dmv_others%>%
  filter(!is.na(Income_level))
```
```{r}
anova_one_way <- aov(Low_Response_Score ~ Income_level, data = dmv_others)
summary(anova_one_way)
```


### Household receiving public assistance
```{r, figsize = 20, figwidth = 8}
dmv_others %>%
  ggplot(mapping = aes(x = pct_PUB_ASST_INC_ACS_13_17, y = Low_Response_Score), scale = x_0
         ) +
  geom_point() +
  geom_smooth() 
```

```{r, figwidth = 10}
dmv_others %>%
  ggplot(mapping = aes(x = pct_PUB_ASST_INC_ACS_13_17, y = Low_Response_Score)) +
  geom_point(size = 0.5, alpha = 0.5) +
  xlim(c(0,20)) +
  geom_smooth(span = 0.2) 
```

###############################################################################################################################################

## Health Insurance
```{r}
insur <- dmv_others[c('State_name', 'pct_One_Health_Ins_ACS_13_17', 'pct_TwoPHealthIns_ACS_13_17', 'pct_No_Health_Ins_ACS_13_17', 'Low_Response_Score')]

insur_ref <- insur %>%
  gather(pct_One_Health_Ins_ACS_13_17:pct_No_Health_Ins_ACS_13_17, key = 'Insurance', value = 'Value')
```

```{r, fig.width = 15}
insur_ref %>%
  ggplot(mapping = aes(x = Value, y = Low_Response_Score, color = Insurance)) +
  geom_point() +
  facet_grid(.~State_name)
```




















