---
author: "Liying Lu"
title: "Tract Data Exploration"
output: html_notebook
---
```{r}
library(dplyr)
library(tidyverse)
library(factoextra)
library(ggplot2)
library(corrplot)
library(ggpubr)
```


## What advice would you give local governments who wish to improve the response to the 2020 census?
```{r}
# import data
tractData <- read.csv("~/PersonalProject/DataFestPrep/data/2019PlanningDatabaseTractData/pdb2019trv6_us.csv")
```
```{r}
names(tractData)
```

Get a smaller dataset by subsetting only the data in DMV region. D.C. Maryland and Virginia. 
```{r}
tractDMV <- tractData %>%
  filter(State_name=="Virginia" |
           State_name=="Maryland" | 
           State_name=="District of Columbia")

head(tractDMV)
```

Summary statistics of the low_response_score on each state.
```{r}
print(summary(tractDMV$Low_Response_Score))
maryland <- filter(tractDMV, State_name=="Maryland")
print(summary(maryland$Low_Response_Score))
virginia <- filter(tractDMV, State_name=="Virginia")
print(summary(virginia$Low_Response_Score))
dc <- filter(tractDMV, State_name=="District of Columbia")
print(summary(dc$Low_Response_Score))
```

### Cultural: Race, Language spoken, Citizenship
**Total Population**: Columns 14, 15
**Low Response Rate**: Columns 284
**Race**: Columns 46-66
**Language spoken**: Columns 69-96 and 179-188
**Citizenship**: Columns 145-152
```{r}
# get the cultural dataset for DMV region
dmv_cultural <- tractDMV %>% select(c(3,5,284,(14:15),(46:66),(69:96), (179:188), (145:152)))
dmv_race <- tractDMV %>% select(c(3,5,284,(46:66)))
dmv_lang <- tractDMV %>% select(c(3,5,284,(69:96), (179:188)))
dmv_citizen <- tractDMV %>% select(c(284, (145:152)))
dmv_pop <- tractDMV %>% select((14:15))
```

## Race
```{r}
dmv_race <- dmv_race %>%
  select(-contains("MOE"))
#names(dmv_race)
pop <- dmv_pop$Tot_Population_ACS_13_17
dmv_race_pct <- data.frame(
  state = dmv_race$State_name,
  low_response_rate = dmv_race$Low_Response_Score,         # response rate
  white = dmv_race$NH_White_alone_ACS_13_17 / pop, # white
  black = dmv_race$NH_Blk_alone_ACS_13_17 / pop,   # black or African American
  aian = dmv_race$NH_AIAN_alone_ACS_13_17 / pop,   # American Indian and Alaska Native
  asian = dmv_race$NH_Asian_alone_ACS_13_17 / pop, # Asian
  native_hawaiian_PI = dmv_race$NH_NHOPI_alone_ACS_13_17 / pop, # Native Hawaiian and other Pacific Islander
  other = dmv_race$NH_SOR_alone_ACS_13_17 / pop    # Some other race
  )
unique(dmv_race_pct$state)
```
```{r}
dmv_race_pct <- dmv_race_pct %>%
  gather(key = "race", value = "pct", -c(state,low_response_rate))
dmv_race_pct
```

```{r}
race_stats <- data.frame(summary(dmv_race_pct))
dmv_race_pct %>%
  ggplot(mapping = aes(y = low_response_rate, x = pct, color=race)) +
  geom_point(alpha = 0.35) +
  geom_smooth(color = "black", alpha = 0.8, se=TRUE, method = "lm") +
  facet_wrap(~race, nrow=2, scale = "free_x")
```
```{r, fig.asp=1, fig.width=10}
race_stats <- data.frame(summary(dmv_race_pct))
dmv_race_pct %>%
  ggplot(mapping = aes(y = low_response_rate, x = pct, color=race)) +
  geom_point(alpha = 0.35) +
  geom_smooth(color = "black", alpha = 0.8, se=TRUE, method = "lm") +
  facet_wrap(~race*state, nrow=2, scale = "free_x")
```

```{r}
dmv_race_pct %>%
  filter(!(race == "white" | race == "black" )) %>%
  ggplot(mapping = aes(y = low_response_rate, x = pct, color=race)) +
  geom_point(alpha = 0.35) +
  geom_smooth(color = "black", alpha = 0.8, se=TRUE) +
  facet_wrap(~race, nrow=2, scale = "free_x")
```

```{r}
# create a linear model for the black and white race
race_black <- dmv_race_pct %>% filter(race == "black")
lm_black <- lm(low_response_rate ~ pct, data = race_black)
summary(lm_black)
```
```{r}
cor.test(race_black$low_response_rate, race_black$pct)
```
```{r}
lm_black2 <- lm(low_response_rate ~ pct * state, data = race_black)
summary(lm_black2)
```


```{r}
# create a linear model for the black and white race
race_white <- dmv_race_pct %>% filter(race == "white")
lm_white <- lm(low_response_rate ~ pct, data = race_white)
summary(lm_white)
```
```{r}
cor.test(race_white$low_response_rate, race_white$pct)
```
```{r}
lm_white2 <- lm(low_response_rate ~ pct + state, data = race_white)
summary(lm_white2)
```
```{r}
unique(dmv_lang$State_name)
names(dmv_lang)
```


```{r}
# person of age 5 years or older who speak english less than very well who speak ----.  
dmv_lang_pct <- data.frame(
  state = dmv_lang$State_name,
  low_response_rate = dmv_lang$Low_Response_Score,
  english = dmv_lang$Age5p_Only_English_ACS_13_17 / pop,
  spanish = dmv_lang$Age5p_Spanish_ACS_13_17 / pop,
  french = dmv_lang$Age5p_French_ACS_13_17 / pop,
  german = dmv_lang$Age5p_German_ACS_13_17 / pop,
  chinese = dmv_lang$Age5p_Chinese_ACS_13_17 / pop,
  russian = dmv_lang$Age5p_Russian_ACS_13_17 / pop,
  viet = dmv_lang$Age5p_Vietnamese_ACS_13_17 / pop,
  arabic = dmv_lang$Age5p_Arabic_ACS_13_17 / pop,
  korean = dmv_lang$Age5p_Korean_ACS_13_17 /pop,
  other_euro = dmv_lang$Age5p_OthEuro_ACS_13_17 / pop, # such as Romanian
  tagalog = dmv_lang$Age5p_Tagalog_ACS_13_17 / pop, # phillipine
  others = dmv_lang$Age5p_OthUnSp_ACS_13_17 / pop # unspecified
)
```

```{r}
dmv_lang_pct <- dmv_lang_pct %>%
  gather(key= "lang", value="pct", -c(state, low_response_rate))
str(dmv_lang_pct)
head(dmv_lang_pct)
```
```{r}

```



```{r, fig.asp=1, fig.width=10}
dmv_lang_pct %>%
  ggplot(mapping = aes(y=low_response_rate, x=pct,  color=lang)) +
  geom_point(alpha = 0.5) +
  geom_smooth(color = "black", method = "lm") +
  facet_wrap(~lang, nrow = 4, scale = "free_x")
```
```{r, fig.asp=1, fig.width=10}
dmv_lang_pct %>%
  ggplot(mapping = aes(y=low_response_rate, x=pct,  color=lang)) +
  geom_point(alpha = 0.5) +
  geom_smooth(color = "black", method = "lm") +
  facet_wrap(~lang*state, nrow = 4, scale = "free_x")
```

```{r}
dmv_lang_pct %>%
  #filter(lang == "english" | lang == "viet") %>%
  ggplot() +
  geom_boxplot(mapping = aes(y = low_response_rate, x = lang))
```

```{r}
unique(dmv_lang_pct$lang)
```

```{r}
dmv_lang_pct %>%
  filter(!is.na(low_response_rate)) %>%
  group_by(lang) %>%
  summarise(
    median = median(low_response_rate),
    mean = mean(low_response_rate)
  )
```





```{r}
lm_lang <- lm(low_response_ ~ pct, data = dmv_lang_pct)
summary(lm_lang)
```

```{r}
eng <- dmv_lang_pct %>% filter(lang == "english")
lm_eng <- lm(low_response_rate ~ pct, data = eng)
summary(lm_eng)
```




























```{r}
household <- tractData$Tot_Occp_Units_ACS_13_17
# limited english speaking household with --- as the main language for communication in household.
dmv_lang_house_pct <- data.frame(
  state = dmv_lang$State_name,
  response_rate = dmv_lang$Low_Response_Score,
  spanish = dmv_lang$ENG_VW_SPAN_ACS_13_17 / household,
  indo_euro = dmv_lang$ENG_VW_INDO_EURO_ACS_13_17 / household,
  asianPI = dmv_lang$ENG_VW_API_ACS_13_17 / household,
  other = dmv_lang$ENG_VW_OTHER_ACS_13_17 / household,
  no_english = dmv_lang$ENG_VW_ACS_13_17 / household
)
#dmv_lang_house_pct %>%
#  dmv_lang_house_pct$english = 1 - spanish - indo_euro - asianPI - other - no_english
str(dmv_lang_house_pct)
```
```{r}
dmv_lang$ENG_VW_SPAN_ACS_13_17 / household

```





























